A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-10 tex/codatadp.tex
Save space on example, begin discussion of model values.

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In this rule, $\sigma$ is a complete mapping over the equivalence classes that occur as sources in the edges of paths in $\tpath{\Ec}{\ec{t_2}}$.
The rule applies if $\tpath{\Ec}{\ec{t_1}}$ is comprised of the paths $p \sigma$ for each (and only) the paths $p \in \tpath{\Ec}{\ec{t_2}}$,
in which case $t_1$ must be equal to $t_2$.
We demonstrate this rule concretely with the following example.
\rem{ensure that \const{LCons} understood}

\begin{example}
Let $\Ec$ be the set $\{ x \teq \const{LCons}(y,\const{LCons}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{LCons}(y,\const{LCons}(y,x)) \}$, $\{ \const{LCons}(y,x) \}$, and $\{ y \}$.
The set $\tpath{\Ec}{\ec{x}}$ is:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{x} \rightarrow^{\const{LCons}}_1 & \ec{y}, & & \\ 
   & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_1 & \ec{y},& \\
   & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} & \}.
\end{array}\]
The set $\tpath{\Ec}{\ec{\const{LCons}(y,x)}}$ is:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_1 & \ec{y}, & & \\ 
   & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} \rightarrow^{\const{LCons}}_1 & \ec{y},& \\
   & \ec{\const{LCons}(y,x)} \rightarrow^{\const{LCons}}_2 & \ec{x} \rightarrow^{\const{LCons}}_2 & \ec{\const{LCons}(y,x)} & \}.
\end{array}\]
Thus, we may infer $x \teq \const{LCons}(y,x)$ by the rule \rn{Bisimilar}
for the mapping $\sigma :=$ $\{ \ec{x} \mapsto \ec{\const{LCons}(y,x)}, \ec{\const{LCons}(y,x)} \mapsto \ec{x} \}$.
\qed
\end{example}

@@

Given a set of disjoint paths $P$ where each $p \in P$
we write $\interp{P}$ to denote the value

@@

\begin{comment}
\rn{Ord}
\(
\inferrule{
  \ec{x} \not\in \vdom(\Val)
  \quad
  x : \tau
  \quad
  \tau \in \Nondata
  \quad
  c \not\in \vrange(\Val)
}{
  \Val := \Val \cup \ec{x} \mapsto c
}
\)

\rn{D-Ind}
\(
\inferrule{
  \ec{x} \not\in \vdom(\Val)
  \quad
  x : \tau
  \quad
  \tau \in \Data
  \quad
  C( \vec t ) \in \ec{x}
  \quad
  \ec{\vec t} \in \vdom(\Val)
}{
  \Val := \Val \cup \ec{x} \mapsto C( \Val \ec{\vec t}  )
}
\)

\rn{D-Base}
\(
\inferrule{
  \ec{x} \not\in \vdom(\Val)
  \quad
  x : \tau
  \quad
  \tau \in \Data
  \quad
  \ec{x} \text{ ctr-free}
  \quad
  c \not\in \vrange(\Val)
}{
  \Val := \Val \cup \ec{x} \mapsto c
}
\)
\end{comment}

\rn{Ind}
\(
\inferrule{
  %x : \tau
  %\quad
  %\tau \in \Codata
  v_{\ec{x}} \in FV( \Val )
  \quad
  C( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  C \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. C( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\)

\rn{Base}
\(
\inferrule{
  %x : \tau
  %\quad
  %\tau \in \Codata
  %\quad
  v_{\ec{x}} \in FV( \Val )
  %\quad
  %\ec{x} \text{ ctr-free}
  \quad
  c \not\sim c' \text{ for all } c' \in \vrange(\Val)
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto c \}
}
\)

\rem{define $\sim$}
We assume that first \rn{Ind} is applied exhaustively, after which \rn{Base} is applied exhaustively.
%the rule \rn{D-Ind} is applied to completion before applying \rn{D-Base}.
In the rule \rn{Ind}, we write $FV( \Val )$ to denote the free variables occurring in $\Val$.
Notice that this excludes variables beneath $\mu$-binders.

As an example,
say that $\Fc$ contains equivalence classes $\ec{x}$ and $\ec{y}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$, where $z$ and $w$ are variables of integer type.
A possible application of the above rules, showing the state of $\Val$ after each application, is as follows:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\hline
\text{Rule} & \multicolumn{4}{c}{\Val} \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{z}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( \const{0}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{w}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( \const{1}, \mu v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( \const{0}, \mu v_{\ec{x}}. \const{C}( \const{1}, v_{\ec{y}} ) ), & \ec{w} \mapsto \const{1} & \} \\
\end{array}\]

It is clear that after applying these rules to completion, $\Val$ is a mapping from equivalence classes to ground values.

>>>>

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In this rule, $\sigma$ is a complete mapping over the equivalence classes that occur as sources in the edges of paths in $\tpath{\Ec}{\ec{t_2}}$.
The rule applies if $\tpath{\Ec}{\ec{t_1}}$ is comprised of the paths $p \sigma$ for each (and only) the paths $p \in \tpath{\Ec}{\ec{t_2}}$,
in which case $t_1$ must be equal to $t_2$.
We demonstrate an application of this rule concretely with the following example.

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After applying the rules in Figure~\ref{fig:cc-rules} to completion, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
The sets $\tpath{\Ec}{\ec{x}}$ and $\tpath{\Ec}{\ec{\const{C}(y,x)}}$ are:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\{ & \ec{x} \rightarrow^{\const{C}}_1 & \ec{y}, & & & 
\{ & \ec{\const{C}(y,x)} \rightarrow^{\const{C}}_1 & \ec{y}, & & \\ 
   & \ec{x} \rightarrow^{\const{C}}_2 & \ec{\const{C}(y,x)} \rightarrow^{\const{C}}_1 & \ec{y},&  &
   & \ec{\const{C}(y,x)} \rightarrow^{\const{C}}_2 & \ec{x} \rightarrow^{\const{C}}_1 & \ec{y},& \\
   & \ec{x} \rightarrow^{\const{C}}_2 & \ec{\const{C}(y,x)} \rightarrow^{\const{C}}_2 & \ec{x} & \} &
   & \ec{\const{C}(y,x)} \rightarrow^{\const{C}}_2 & \ec{x} \rightarrow^{\const{C}}_2 & \ec{\const{C}(y,x)} & \}.
\end{array}\]
Thus, by the rule \rn{Bisimilar}
for the mapping $\sigma :=$ $\{ \ec{x} \mapsto \ec{\const{C}(y,x)}, \ec{\const{C}(y,x)} \mapsto \ec{x} \}$,
we may infer $x \teq \const{C}(y,x)$.
\qed
\end{example}

@@

%Given a set of disjoint paths $P$ where each $p \in P$
%we write $\interp{P}$ to denote the value

Recall that the $\mu$-notation can be used to represent (possibly cyclic) values taken from an initial algebra, \rem{elaborate?} 
e.g. $\mu x. \const{C}( \const{0}, x )$ represents the infinite sequence $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
In this setting, a value can be viewed as a term-like structure where all subterms 
(which we will refer to as \emph{subvalues}) are either 
$\mu x. \const{C}( \vec v )$ for some constructor $\const{C}$ and subvalues $\vec v$
\footnote{We assume all such values are well-typed.}, or a variable $x$.
We will refer to values of this form as \emph{$\mu$-values}.
We say a $\mu$-value is \emph{closed} if it contains no free variables.
For convienience, we will represent the values of both datatypes and co-datatypes as closed $\mu$-values,
where the case of datatypes adds the restriction that none of its subvalues are variables.

To reason about the uniqueness of $\mu$-values, we introduce the following notions.
First, the \emph{interpretation of subvalue $u$ in $v$} is the value $v\interp{u}^\emptyset$ returned by the following,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subvalue of $v$ that binds this occurrence of variable $x$.

\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & x & x \in M \\
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{\mu x. C( \vec t )}^M & = & x & x \in M \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M
\end{array}\]

We say a (closed) $\mu$-value $v$ of the form $\mu x. \const{C}( \vec w )$ is \emph{normal} if and only if 
$v\interp{u}^\emptyset$ is not equivalent to $v$ modulo renaming for all of its strict subvalues $u$,
and all of its strict closed subvalues are normal.
For instance, the value $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is equivalent to $v$ modulo the renaming $\{ x \mapsto y, y \mapsto x \}$.

@@

\rn{Ind}
\(
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  C( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  C \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. C( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\)

\rn{Base}
\(
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  c \not\sim c' \text{ for all } c' \in \vrange(\Val)
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto c \}
}
\)

\rem{define $\sim$}

In the rule \rn{Ind}, we write $FV( \Val )$ to denote the free variables occurring in $\Val$.
Notice that this excludes variables beneath $\mu$-binders.
As an example,
say that $\Fc$ contains equivalence classes $\ec{x}$ and $\ec{y}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$, where $z$ and $w$ are variables of integer type.
A possible application of the above rules, showing the state of $\Val$ after each application, is as follows:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\hline
\text{Rule} & \multicolumn{4}{c}{\Val} \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Ind} \ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{z}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( \const{0}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\rn{Base} \ec{w}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( \const{1}, \mu v_{\ec{y}}. \const{C}( \const{0}, v_{\ec{x}} ) ), & \ec{z} \mapsto \const{0}, \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( \const{0}, \mu v_{\ec{x}}. \const{C}( \const{1}, v_{\ec{y}} ) ), & \ec{w} \mapsto \const{1} & \} \\
\end{array}\]

In our construction, the rule \rn{Ind} is first applied exhaustively.
Since the height of any value mapped to by $\Val$ is bounded by the number of equivalence classes of $\Fc$,
the rule \rn{Ind} can be applied only a finite number of times.
Second, \rn{Base} is applied exhaustively.
Since \rn{Base} removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
the rule \rn{Base} can only be applied a finite number of times and does not trigger any new applications of \rn{Ind}.

We first argue that $\Val$ contains only ground values.
First, since \rn{Split} does not apply to $\Fc$,
all equivalence classes of finite type contain constructor terms.
Thus, after applying \rn{Ind} exhaustively, the free variables of $\Val$ all have infinite type.

%It is clear that after applying these rules to completion, $\Val$ is a mapping from equivalence classes to ground values.
